Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-23 manu-draft.Rmd
methods draft done

@@

* Historical timeline: https://docs.google.com/document/d/1KqaEAYEG7pdcGpWbUjONfVKMHHB6_kOkKCyncp4JYKA/edit, https://thebradentontimes.com/piney-point-a-retrospective-p6328-158.htm
* Initial response, data collection
* Synthesis and Analysis

Need to note that K. brevis cell data are from event-based sampling, i.e., not appropriate to combine with other data for stats analysis.

All stats based on FLDEP monitoring, fixed sample design, etc.

For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.

@@

```{r trnfrq, fig.cap = 'Frquency occurrence estimates for (a) area 1 and (b) area 2 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = `r nrow(rstrnpts)`) near Piney Point.  Estimates are grouped by sample months in 2021. Frequency occurrences are absolute for each taxa based on presence/absence.'}
knitr::include_graphics(here('figs/trnfrq.jpeg'))
```

@@

```{r wqcmps}
cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.  Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
knitr::kable(wqcmptab, booktabs = T, caption = cap.val)
```

>>>>

Water quality data in Tampa Bay have been collected at fixed sampling sites since 1974 by the Environmental Protection Commission of Hillsborough County.  These data include samples at 45 stations used by the TBEP to assess annual progress towards programmatic goals and regulatory thresholds applicable for each of the major segments in Tampa Bay.  Discrete water samples are collected monthly at mid-depth and processed in the laboratory immediately after collection. Similarly, the Parks and Natural Resources Department of Manatee County have been collecting data in Tampa Bay at locations south of Piney Point and south of the main dredged channel of Tampa Bay (approximate longitudinal axis of the bay).  The sampling design is similar to the EPC data, with the exception that sites are sampled approximately every three months but at a higher spatial density per unit area.  Additionally, the period of record for monitoring data from Manatee County began in 1996.

Long-term water quality monitoring data from Hillsborough and Manatee counties were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.  Station data from Hillsborough County were obtained for the middle and lower segments of Tampa Bay.  Station data from Manatee County were obtained for areas in lower Tampa Bay, Terra Ceia Bay, Anna Maria Sound, and northern Sarasota Bay.  For the same water quality parameters noted above (i.e., nitrogen, phosphorus, pH, temperature, salinity), observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference concentrations relative to results at the closest monitoring station that was sampled in response to Piney Point.  This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in August).  Spatial matching of each response monitoring station relative to the long-term monitoring stations was accomplished using the "st_nearest_feature()" function from the sf R package [@Pebesma18].  In some cases, the nearest long-term station did not include data for a every monitoring parameter at a response location and the next closest station was used as reference. Long-term water quality data are available from the University of South Florida Water Atlas (https://wateratlas.usf.edu/).

@@

The R statistical programming language (v4.0.2) was used to import, synthesize, and analyze all datasets provided by multiple partners [@RCT20].  Partner data were uploaded or entered manually as Google spreadsheets, where they were imported into R using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages.  The suite of R packages available in the tidyverse [@Wickham19] were used to wrangle the data into an appropriate format for analysis.  The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data for Tampa Bay, specifically the EPC water quality data and seagrass transect database.

Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparisons tests to assess trends over time.  Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general assessment of potential patterns in eutrophication following wastewater release. Further analyses were conducted to compare all data types together, including the entire suite of water quality data, seagrass frequency occurrences, and macroalgae frequency occurrences to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.  Data describing *K. brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.  Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.

The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results. For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.

@@

```{r trnfrq, fig.cap = 'Frequency occurrence estimates for (a) area 1 and (b) area 2 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = `r nrow(rstrnpts)`) near Piney Point.  Estimates are grouped by sample months in 2021. Frequency occurrences are absolute for each taxa based on presence/absence.'}
knitr::include_graphics(here('figs/trnfrq.jpeg'))
```

@@

```{r wqcmptab}
cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.  Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
knitr::kable(wqcmptab, booktabs = T, caption = cap.val)
```
